Determinism, Complexity, and Predictability in Computer Performance

### 2014-07-31 results.tex
swap two plots, update text to reflect

Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. MASE plots for the full
set of experiments.
\begin{figure*}
  \centering
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/LMA_prediction_vs_entropy}
  \includegraphics[width=1.7\columnwidth]{figs/new_predictions_vs_entropy4a}
\caption{WPE vs. MASE for all trials, methods, and systems---with the
  exception of \svdone, \svdthree, and \svdfive are omitted from the
  top-left plot for scale reasons, as described in the text.
%
% ; their MASE scores are $2.676 \pm 4.328$, $31.39 \pm 0.282$, and
%   $20.87 \pm 0.192$ respectively.
%
Numerical values, including means and standard deviations of the
errors, can be found in Table~\ref{tab:error}.  The curves and shaded
regions are the same as in the previous figure.  }
    \label{fig:wpe_vs_mase_all}
\end{figure*}
There are 15 points in each cluster, one for each trial.  (The points
in Figure~\ref{fig:wpe_vs_mase_best} are the leftmost of the points
for the corresponding trace in any of the four plots in
Figure~\ref{fig:wpe_vs_mase_all}.)  The WPE values do not vary very
much across trials.  For most traces, the variance in MASE scores is
low as well, resulting in small, tight clusters.  In some
cases---\arima predictions of {\tt col\_major}, for instance---the
MASE variance is larger, which spreads out the clusters horizontally.
The mean MASE scores of predictions generated with the nonlinear LMA
method, for instance, are generally closer to the dashed curve; the
\arima method clusters are more widely spread, the \naive ~clusters
even more so.  A few of the clusters have very high variance; these
are discussed later in this section.

The main thing to note here, however, is not the details of the shapes
of the clusters, but rather their positions in the four plots:
specifically, the fact that many of them are to the right of and/or
below the dashed curve that identifies the boundary of the shaded
region.  \emph{These predictions are not as good as our heuristic
  suggests they could be.}  Focusing in on any single cluster of
points makes this clear: LMA works best for \svdsix, for instance,
followed by the random-walk prediction method, then \arima and \naive.
When a practioner calculates an WPE vs. MASE value that is outside the
shaded region, that should suggest to her that the prediction method
is not well matched to the task at hand: that is, the time series has
more predictive structure than the method is able to use.
%
% The idea behind the shaded region is to help practitioners identify
% when there is predictive structure in the data that a given method is
% not exploiting (viz., \arima on \svdsix)---i.e., that one could do
% better with another forecast strategy.
%
In this case---e.g., \arima on \svdsix---she should try a different
method.  The position of the LMA cluster for \svdsix, on the other
hand, reflects this method's ability to capture and exploit the
structure that is present in this signal.  WPE vs. MASE values like
this, which fall in the shaded region, should suggest to the
practitioner that the prediction method is well-suited to the task.
The following discussion lays out the details that underlie these
claims.
%
% The dashed curve in Figures~\ref{fig:wpe_vs_mase_best}
% and~\ref{fig:wpe_vs_mase_all} is a heuristic that captures the
% argument at the end of the previous paragraph.
%

@@

This disparity---WPE values that suggest a high rate of forward
information transfer in the signal, but predictions with comparatively
poor MASE scores---is obvious in the geometry of the two left-hand
images in Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters
are far to the right of and/or below the dashed curve.  Again, this
indicates that these methods are not leveraging the available
information in the signal.  The dynamics of \col may be complicated,
but they are not unstructured.  This signal is nonlinear and
deterministic~\cite{mytkowicz09}, and if one uses a prediction
technique that is based a nonlinear model (LMA)---rather than a method
that simply predicts the running mean (\naive) or the previous value
(random walk), or one that uses a linear model (\arima)---the MASE
score is much better: $0.050 \pm 0.001$.  This prediction is 20 times
more accurate than a random-walk forecast, which is more in line with
the level of predictive structure that the low WPE value suggests is
present in the signal.  The MASE scores of random-walk predictions of
\col are all $\approx 1$---as one would expect---pushing those points
well below the shaded region.  Clearly the stationarity assumption on
which this method is based does not hold for this signal.

@@

\svdone---the dark blue segment of
Figure~\ref{fig:svd-ts-colored}---behaves very differently than the
other seven systems in this study.  Though its weighted permutation
entropy is very high ($0.957 \pm 0.016$), three of the four prediction
methods do quite well on this signal, yielding mean MASE scores of
0.714 (\arima), 0.827 (LMA), and 0.933 (random walk).  This pushes the
corresponding clusters of points in Figure~\ref{fig:wpe_vs_mase_all}
well above the trend followed by the other seven signals.  The reasons
for this are discussed in the following paragraph.  The MASE scores of
the predictions that were produced by the \naive ~method for this
system, however, are highly inconsistent.  The majority of the blue
diamond-shaped points on the top-left plot in
Figure~\ref{fig:wpe_vs_mase_all} are clustered near a MASE score of
0.6, which is better than the other three methods.  In five of the 15
\svdone trials, however, there were step changes in the signal.  The
\naive ~method has a very difficult time with signals like this,
particularly if there are multiple step changes.  This raised the MASE
scores of these trials, pushing the corresponding points to the
right\footnote{This includes the cluster of three points near MASE
  $\approx 2.5$, as well as two points that are beyond the domain of
  the graph, at MASE $\approx 11.2-14.8$.}, and in turn raising both
the mean and variance of this set of trials.

@@

Again, the curves in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} were determined from a finite set of
methods and data.  We put a lot of thought and effort into making
these experiments upon which these results are based both
representative and comprehensive.  The forecast methods involved range
from the simple to the sophisticated; the time-series data was sampled
from a system whose behavior spans the dynamical behavior space.
While we are cautiously optimistic about the generality of our
conclusions, more exploration will be required before we can make any
such claim.  Our preliminary work along those lines shows that data
from the Henon map \cite{henon}, the Lorenz system \cite{lorenz}, and
the SFI A data set \cite{sfi-data} all fall within the one-$\sigma$
volume of the fit in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} region, as do various nonlinear
transformations of \svdtwo and \svdsix.

>>>>

Figure~\ref{fig:wpe_vs_mase_all} shows WPE vs. MASE plots for the full
set of experiments.
\begin{figure*}
  \centering
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/ARIMA_prediction_vs_entropy}
%    \includegraphics[width=\columnwidth]{figs/LMA_prediction_vs_entropy}
  \includegraphics[width=1.7\columnwidth]{figs/new_predictions_vs_entropy4a}
\caption{WPE vs. MASE for all trials, methods, and systems---with the
  exception of \svdone, \svdthree, and \svdfive are omitted from the
  top-right plot for scale reasons, as described in the text.
%
% ; their MASE scores are $2.676 \pm 4.328$, $31.39 \pm 0.282$, and
%   $20.87 \pm 0.192$ respectively.
%
Numerical values, including means and standard deviations of the
errors, can be found in Table~\ref{tab:error}.  The curves and shaded
regions are the same as in the previous figure.  }
    \label{fig:wpe_vs_mase_all}
\end{figure*}
There are 15 points in each cluster, one for each trial.  (The points
in Figure~\ref{fig:wpe_vs_mase_best} are the leftmost of the points
for the corresponding trace in any of the four plots in
Figure~\ref{fig:wpe_vs_mase_all}.)  The WPE values do not vary very
much across trials.  For most traces, the variance in MASE scores is
low as well, resulting in small, tight clusters.  In some
cases---\arima predictions of {\tt col\_major}, for instance---the
MASE variance is larger, which spreads out the clusters horizontally.
The mean MASE scores of predictions generated with the nonlinear LMA
method, for instance, are generally closer to the dashed curve; the
\arima method clusters are more widely spread, the \naive ~clusters
even more so.  A few of the clusters have very high variance; these
are discussed later in this section.

The main thing to note here, however, is not the details of the shapes
of the clusters, but rather their positions in the four plots:
specifically, the fact that many of them are to the right of and/or
below the dashed curve that identifies the boundary of the shaded
region.  \emph{These predictions are not as good as our heuristic
  suggests they could be.}  Focusing in on any single cluster of
points makes this clear: LMA works best for \svdsix, for instance,
followed by the random-walk prediction method, then \arima and \naive.
When a practitioner calculates an WPE vs. MASE value that is outside the
shaded region, that should suggest to her that the prediction method
is not well matched to the task at hand: that is, the time series has
more predictive structure than the method is able to use.
%
% The idea behind the shaded region is to help practitioners identify
% when there is predictive structure in the data that a given method is
% not exploiting (viz., \arima on \svdsix)---i.e., that one could do
% better with another forecast strategy.
%
In this case---e.g., \arima on \svdsix---she should try a different
method.  The position of the LMA cluster for \svdsix, on the other
hand, reflects this method's ability to capture and exploit the
structure that is present in this signal.  WPE vs. MASE values like
this, which fall in the shaded region, should suggest to the
practitioner that the prediction method is well-suited to the task.
The following discussion lays out the details that underlie these
claims.
%
% The dashed curve in Figures~\ref{fig:wpe_vs_mase_best}
% and~\ref{fig:wpe_vs_mase_all} is a heuristic that captures the
% argument at the end of the previous paragraph.
%

@@

This disparity---WPE values that suggest a high rate of forward
information transfer in the signal, but predictions with comparatively
poor MASE scores---is obvious in the geometry of the two left-hand
and the upper-right hand images in Figure~\ref{fig:wpe_vs_mase_all}, where the \col clusters
are far to the right of and/or below the dashed curve.  Again, this
indicates that these methods are not leveraging the available
information in the signal.  The dynamics of \col may be complicated,
but they are not unstructured.  This signal is nonlinear and
deterministic~\cite{mytkowicz09}, and if one uses a prediction
technique that is based a nonlinear model (LMA)---rather than a method
that simply predicts the running mean (\naive) or the previous value
(random walk), or one that uses a linear model (\arima)---the MASE
score is much better: $0.050 \pm 0.001$.  This prediction is 20 times
more accurate than a random-walk forecast, which is more in line with
the level of predictive structure that the low WPE value suggests is
present in the signal.  The MASE scores of random-walk predictions of
\col are all $\approx 1$---as one would expect---pushing those points
well below the shaded region.  Clearly the stationarity assumption on
which this method is based does not hold for this signal.

@@

\svdone---the dark blue segment of
Figure~\ref{fig:svd-ts-colored}---behaves very differently than the
other seven systems in this study.  Though its weighted permutation
entropy is very high ($0.957 \pm 0.016$), three of the four prediction
methods do quite well on this signal, yielding mean MASE scores of
0.714 (\arima), 0.827 (LMA), and 0.933 (random walk).  This pushes the
corresponding clusters of points in Figure~\ref{fig:wpe_vs_mase_all}
well above the trend followed by the other seven signals.  The reasons
for this are discussed in the following paragraph.  The MASE scores of
the predictions that were produced by the \naive ~method for this
system, however, are highly inconsistent.  The majority of the blue
diamond-shaped points on the top-right plot in
Figure~\ref{fig:wpe_vs_mase_all} are clustered near a MASE score of
0.6, which is better than the other three methods.  In five of the 15
\svdone trials, however, there were step changes in the signal.  The
\naive ~method has a very difficult time with signals like this,
particularly if there are multiple step changes.  This raised the MASE
scores of these trials, pushing the corresponding points to the
right\footnote{This includes the cluster of three points near MASE
  $\approx 2.5$, as well as two points that are beyond the domain of
  the graph, at MASE $\approx 11.2-14.8$.}, and in turn raising both
the mean and variance of this set of trials.

@@

Again, the curves in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} were determined from a finite set of
methods and data.  We put a lot of thought and effort into making
these experiments upon which these results are based both
representative and comprehensive.  The forecast methods involved range
from the simple to the sophisticated; the time-series data was sampled
from a system whose behavior spans the dynamical behavior space.
While we are cautiously optimistic about the generality of our
conclusions, more exploration will be required before we can make any
such claim.  Our preliminary work along those lines shows that data
from the Henon map \cite{henon}, the Lorenz system \cite{lorenz}, and
the SFI A data set \cite{sfi-data} all fall within the one-$\sigma$
volume of the fit in Figures~\ref{fig:wpe_vs_mase_best}
and~\ref{fig:wpe_vs_mase_all} region, as do various nonlinear
transformations of \svdtwo, \svdfive and \svdsix.
